---
title: "Data Analysis"
author: "Alice Mao, Leila Yan"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message=FALSE)
knitr::opts_chunk$set(warning=FALSE)
```

```{r, include=FALSE}
library(tidyverse)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

```{r, include=FALSE}
# Load required libraries
library(tidyverse)
library(ggplot2)
library(broom)

# Load the dataset
hiv_data <- read.csv("data/HIV_AIDS_Diagnoses_by_Neighborhood__Sex__and_Race_Ethnicity_20241118.csv")

UHF <- read.csv("data/UHF_Neighborhoods.csv")

```

---------------------------------

## How have HIV and AIDS diagnoses changed over time, and do specific neighborhoods experience disproportionately high rates?

---------------------------------

### Analysis

```{r, include=FALSE}
# Clean and prepare the dataset
hiv_data <- hiv_data %>%
  mutate(
    YEAR = as.numeric(YEAR),
    Neighborhood = as.character(Neighborhood..U.H.F.), # Correct column for neighborhood
    HIV_Diagnoses = as.numeric(HIV.DIAGNOSES.PER.100.000.POPULATION),
    AIDS_Diagnoses = as.numeric(AIDS.DIAGNOSES.PER.100.000.POPULATION)
  ) %>%
  filter(YEAR >= 2016, !is.na(HIV_Diagnoses), !is.na(AIDS_Diagnoses)) # Keep only data from 2016 onwards

```

```{r, message=FALSE}
# Aggregate HIV and AIDS diagnoses
hiv_trends <- hiv_data %>%
  group_by(YEAR, Neighborhood) %>%
  summarise(
    avg_hiv_rate = mean(HIV_Diagnoses, na.rm = TRUE),
    avg_aids_rate = mean(AIDS_Diagnoses, na.rm = TRUE)
  ) %>%
  ungroup()

# Preview the aggregated data
head(hiv_trends)
```

```{r, message=FALSE}
# Plot HIV diagnosis trends by neighborhood
ggplot(hiv_trends, aes(x = YEAR, y = avg_hiv_rate, group = Neighborhood, color = Neighborhood)) +
  geom_line(size = 0.5, alpha = 0.6) + 
  labs(
    title = "HIV Diagnoses by Neighborhood",
    x = "Year",
    y = "Average HIV Diagnoses per 100,000 Population",
    color = "Neighborhood"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    legend.text = element_text(size = 5), 
    legend.title = element_text(size = 7),
    plot.title = element_text(hjust = 0.6, size = 9), 
    axis.text = element_text(size = 6),  
    axis.title = element_text(size = 8)  
  )


```

```{r, message=FALSE}
# Calculate the mean HIV diagnosis rate for each neighborhood
top_neighborhoods <- hiv_data %>%
  group_by(Neighborhood) %>%
  summarise(mean_hiv_rate = mean(HIV_Diagnoses, na.rm = TRUE)) %>%
  arrange(desc(mean_hiv_rate)) %>%
  slice(1:10) %>%
  pull(Neighborhood)

# Filter the dataset for the top 10 neighborhoods
filtered_hiv_data <- hiv_data %>%
  filter(Neighborhood %in% top_neighborhoods)

# Aggregate HIV diagnoses for the top 10 neighborhoods
top_hiv_trends <- filtered_hiv_data %>%
  group_by(YEAR, Neighborhood) %>%
  summarise(
    avg_hiv_rate = mean(HIV_Diagnoses, na.rm = TRUE)
  ) %>%
  ungroup()

# Preview the aggregated data
head(top_hiv_trends)

# Plot HIV diagnosis trends for the top 10 neighborhoods
ggplot(top_hiv_trends, aes(x = YEAR, y = avg_hiv_rate, group = Neighborhood, color = Neighborhood)) +
  geom_line(size = 1.2) +
  labs(
    title = "HIV Diagnoses for Top 10 Neighborhoods",
    x = "Year",
    y = "Average HIV Diagnoses per 100,000 Population",
    color = "Neighborhood"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    legend.text = element_text(size = 6), 
    legend.title = element_text(size = 8),
    plot.title = element_text(hjust = 0.5, size = 10), 
    axis.text = element_text(size = 6),  
    axis.title = element_text(size = 8)  
  )

```

```{r, message=FALSE}
# Plot AIDS diagnosis trends by neighborhood with a legend
ggplot(hiv_trends, aes(x = YEAR, y = avg_aids_rate, group = Neighborhood, color = Neighborhood)) +
  geom_line(size = 0.8, alpha = 0.7) +
  labs(
    title = "AIDS Diagnoses by Neighborhood",
    x = "Year",
    y = "Average AIDS Diagnoses per 100,000 Population",
    color = "Neighborhood"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    legend.text = element_text(size = 5), 
    legend.title = element_text(size = 7),
    plot.title = element_text(hjust = 0.6, size = 9), 
    axis.text = element_text(size = 6),  
    axis.title = element_text(size = 8)  
  )

```

```{r}
# Calculate mean rates for each neighborhood
top_hiv_neighborhoods <- hiv_data %>%
  group_by(Neighborhood) %>%
  summarise(mean_hiv_rate = mean(HIV_Diagnoses, na.rm = TRUE)) %>%
  arrange(desc(mean_hiv_rate)) %>%
  slice(1:10) %>%
  pull(Neighborhood)

top_aids_neighborhoods <- hiv_data %>%
  group_by(Neighborhood) %>%
  summarise(mean_aids_rate = mean(AIDS_Diagnoses, na.rm = TRUE)) %>%
  arrange(desc(mean_aids_rate)) %>%
  slice(1:10) %>%
  pull(Neighborhood)

# Filter the dataset for top 10 neighborhoods
filtered_hiv_trends <- hiv_data %>%
  filter(Neighborhood %in% top_hiv_neighborhoods) %>%
  group_by(YEAR, Neighborhood) %>%
  summarise(avg_hiv_rate = mean(HIV_Diagnoses, na.rm = TRUE)) %>%
  ungroup()

filtered_aids_trends <- hiv_data %>%
  filter(Neighborhood %in% top_aids_neighborhoods) %>%
  group_by(YEAR, Neighborhood) %>%
  summarise(avg_aids_rate = mean(AIDS_Diagnoses, na.rm = TRUE)) %>%
  ungroup()

# Plot HIV diagnosis trends
ggplot(filtered_hiv_trends, aes(x = YEAR, y = avg_hiv_rate, group = Neighborhood, color = Neighborhood)) +
  geom_line(size = 1) +
  labs(
    title = "HIV Diagnoses in Top 10 Neighborhoods",
    x = "Year",
    y = "Average HIV Diagnoses per 100,000 Population",
    color = "Neighborhood"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    legend.text = element_text(size = 6), 
    legend.title = element_text(size = 8),
    plot.title = element_text(hjust = 0.5, size = 10), 
    axis.text = element_text(size = 6),  
    axis.title = element_text(size = 8)  
  )
```

```{r}
# Fit a linear regression model for year predicting HIV diagnosis rates
hiv_lm <- lm(avg_hiv_rate ~ YEAR, data = hiv_trends)

# Summarize the regression model
hiv_lm_summary <- summary(hiv_lm)

# Display regression coefficients
tidy(hiv_lm) %>%
  select(term, estimate, std.error, p.value) %>%
  mutate(p.value = format(p.value, scientific = TRUE)) %>%  # Show exact p-value in scientific notation
  knitr::kable(digits = 10)

# Identify neighborhoods with consistently high HIV diagnosis rates
high_hiv_neighborhoods <- hiv_trends %>%
  group_by(Neighborhood) %>%
  summarise(mean_hiv_rate = mean(avg_hiv_rate, na.rm = TRUE)) %>%
  arrange(desc(mean_hiv_rate)) %>%
  filter(mean_hiv_rate > quantile(mean_hiv_rate, 0.9)) # Top 10% neighborhoods

# Display the neighborhoods with high HIV rates
high_hiv_neighborhoods %>%
  knitr::kable(digits = 2)

```

```{r}

tidy(hiv_lm) %>%
  select(term, estimate, std.error, p.value) %>%
  mutate(p.value = format(p.value, scientific = TRUE)) %>%  # Show exact p-value in scientific notation
  knitr::kable(digits = 10)

```

---------------------------------

### Conclusion

The analysis of temporal trends in HIV diagnoses demonstrates a significant downward trend in diagnosis rates from 2016 onward. The linear regression model revealed a slope of -1.982, indicating an average annual decrease of approximately 1.98 HIV diagnoses per 100,000 population. The p-value for the `YEAR` variable was lower than 0.05, indicating a statistically significant evidence against the null hypothesis and confirming that the observed trend is not due to random variation. However, neighborhood-level disparities remain a significant concern. Analysis of outlier neighborhoods highlighted that certain areas consistently reported significantly higher HIV diagnosis rates.

---------------------------------

## Are there significant disparities in HIV/AIDS diagnoses across sex and race/ethnicity?

---------------------------------

### Analysis

```{r}
# Group by sex and calculate average HIV diagnoses
hiv_by_sex <- hiv_data %>%
  group_by(SEX) %>%
  summarise(avg_hiv_rate = mean(HIV_Diagnoses, na.rm = TRUE)) %>%
  arrange(desc(avg_hiv_rate))

# Group by race/ethnicity and calculate average HIV diagnoses
hiv_by_race <- hiv_data %>%
  group_by(RACE.ETHNICITY) %>%
  summarise(avg_hiv_rate = mean(HIV_Diagnoses, na.rm = TRUE)) %>%
  arrange(desc(avg_hiv_rate))

# Display the summaries
hiv_by_sex %>%
  knitr::kable(digits = 2, caption = "Average HIV Diagnosis Rates by Sex")

hiv_by_race %>%
  knitr::kable(digits = 2, caption = "Average HIV Diagnosis Rates by Race/Ethnicity")

# Bar plot for HIV diagnosis rates by sex
ggplot(hiv_by_sex, aes(x = SEX, y = avg_hiv_rate, fill = SEX)) +
  geom_bar(stat = "identity") +
  labs(
    title = "HIV Diagnosis Rates by Sex",
    x = "Sex",
    y = "Average HIV Diagnoses per 100,000 Population"
  ) +
  theme_minimal()

# Bar plot for HIV diagnosis rates by race/ethnicity
ggplot(hiv_by_race, aes(x = reorder(RACE.ETHNICITY, avg_hiv_rate), y = avg_hiv_rate, fill = RACE.ETHNICITY)) +
  geom_bar(stat = "identity") +
  coord_flip() + # Flip for better readability
  labs(
    title = "HIV Diagnosis Rates by Race/Ethnicity",
    x = "Race/Ethnicity",
    y = "Average HIV Diagnoses per 100,000 Population"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```

```{r}
# Fit a linear regression model for HIV diagnosis rates
hiv_lm_2 <- lm(HIV_Diagnoses ~ SEX + RACE.ETHNICITY, data = hiv_data)

# Summarize the regression model
hiv_lm_summary_2 <- summary(hiv_lm_2)

# Display regression coefficients
tidy(hiv_lm_2) %>%
  select(term, estimate, std.error, p.value) %>%
  mutate(p.value = format(p.value, scientific = TRUE)) %>%
  knitr::kable(digits = 3, caption = "Linear Regression: Predicting HIV Diagnoses")


```

```{r}
# Fit a multivariate regression model adjusting for year and neighborhood
hiv_multivariate_lm <- lm(HIV_Diagnoses ~ SEX + RACE.ETHNICITY + YEAR + Neighborhood..U.H.F., data = hiv_data)

# Summarize the multivariate regression model
tidy(hiv_multivariate_lm) %>%
  select(term, estimate, std.error, p.value) %>%
  mutate(p.value = format(p.value, scientific = TRUE)) %>%
  knitr::kable(digits = 10, caption = "Multivariate Regression: Adjusting for Confounders")


```

---------------------------------

### Conclusion

The analysis of demographic disparities in HIV diagnoses revealed significant differences across sex and race/ethnicity. The linear regression model identified males as having significantly higher HIV diagnosis rates than females, with a coefficient of 16.771 for `SEXMale`. The p-value for this variable was lower than 0.05, providing strong statistical evidence that sex is a significant predictor of HIV diagnoses.

Race and ethnicity were also found to be highly significant predictors. The coefficients for `RACE.ETHNICITYBlack` (33.035) and `RACE.ETHNICITYLatino/Hispanic` (6.513) highlight disproportionately high HIV diagnosis rates among Black and Hispanic populations. The near-zero p-values for these predictors confirm that the observed disparities are not due to random variation. These findings align with existing evidence that systemic inequities, including barriers to healthcare access, economic instability, and stigma, disproportionately affect these groups.

The multivariate regression model, adjusting for confounders such as year and neighborhood, confirmed the robustness of these disparities. Even after controlling for these factors, the p-values for sex and race/ethnicity remained small, showing that these disparities are deeply rooted in structural and systemic inequities rather than purely geographic or temporal differences.

---------------------------------

## What is the proportion of concurrent HIV/AIDS diagnoses among total HIV diagnoses, and how does this vary by neighborhood and demographic group?

---------------------------------

```{r, include=FALSE}
# load the package
library(dplyr)
library(ggplot2)
library(sf)
library(RColorBrewer)

# import dataset
Merged_HIV_UHF_Data <- read.csv("Merged_HIV_UHF_Data.csv")
```

```{r}
# create summary data for concurrent HIV/AIDS diagnosis
summary_data <- Merged_HIV_UHF_Data %>%
  group_by(uhf_label, sex, race_ethnicity) %>%
  summarize(
    total_concurrent_hiv_aids = sum(total_number_of_concurrent_hiv_aids_diagnoses, na.rm = TRUE),
    total_hiv_diagnoses = sum(total_number_of_hiv_diagnoses, na.rm = TRUE),
    proportion_concurrent = total_concurrent_hiv_aids / total_hiv_diagnoses
  ) %>%
  arrange(desc(proportion_concurrent))
```

```{r}
# Fit the linear regression model
lm_model <- lm(proportion_concurrent ~ uhf_label + sex + race_ethnicity, data = summary_data)

# Summary of the linear regression model
summary(lm_model)
```

---------------------------------

### Model Discussion

The model explains a small proportion of the variability in the data (R-squared = 0.04472), suggesting that other unmeasured factors may play a more significant role. While the model is statistically significant overall (p = 0.001198), the limited explanatory power highlights the need for additional data or variables to better understand disparities in concurrent diagnoses. These findings suggest that, while neighborhood and most demographic factors do not show strong effects, the elevated proportion observed in the "Other/Unknown" group warrants further investigation into potential disparities in healthcare access or reporting.

---------------------------------

```{r}
coeff_table <- summary(lm_model)$coefficients
coeff_table <- as.data.frame(coeff_table)
colnames(coeff_table) <- c("Estimate", "Std. Error", "t value", "Pr(>|t|)")

# Add significance levels
coeff_table$Significance <- cut(coeff_table$`Pr(>|t|)`, 
                                breaks = c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf), 
                                labels = c("***", "**", "*", ".", " "))

# View the coefficient table
print(coeff_table)
```

```{r}
ggplot(summary_data, aes(x = uhf_label, y = proportion_concurrent, color = race_ethnicity)) +
  geom_point(size = 2, alpha = 0.8) +  # Keep points for visibility
  facet_wrap(~ sex, scales = "free_y") +  # Separate panels for each sex
  labs(
    title = "Proportion of Concurrent HIV/AIDS Diagnoses by Neighborhood and Demographics",
    x = "Neighborhood (UHF Label)",
    y = "Proportion of Concurrent Diagnoses",
    color = "Race/Ethnicity"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5, size = 14),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  ) +
  scale_color_brewer(palette = "Set1")
```

---------------------------------

### Pattern Discussion by Sex and Race/Ethnicity

#### By Sex:

In the "Female" and "Male" categories, the distribution of concurrent diagnoses appears similar in variation but may differ slightly in range and clustering.
Men exhibit a slightly broader range in some neighborhoods compared to women.

#### By Race/Ethnicity:

Black and Latino/Hispanic populations have higher proportions of concurrent diagnoses in several neighborhoods compared to White or Asian/Pacific Islander populations.
Other/Unknown groups show wide variability, likely influenced by limited data or inconsistent categorization.

Certain racial/ethnic groups (e.g., White or Asian/Pacific Islander) appear clustered at lower proportions in many neighborhoods.


```{r}
# neighborhood summary
neighborhood_summary <- Merged_HIV_UHF_Data %>%
  group_by(neighborhood_u_h_f) %>%
  summarize(
    total_concurrent_hiv_aids = sum(total_number_of_concurrent_hiv_aids_diagnoses, na.rm = TRUE),
    total_hiv_diagnoses = sum(total_number_of_hiv_diagnoses, na.rm = TRUE),
    proportion_concurrent = total_concurrent_hiv_aids / total_hiv_diagnoses
  ) %>%
  arrange(desc(proportion_concurrent))

print(neighborhood_summary)
```

```{r}
# visualization of Proportion of Concurrent HIV/AIDS Diagnoses by Neighborhood
ggplot(neighborhood_summary, aes(x = reorder(neighborhood_u_h_f, -proportion_concurrent), y = proportion_concurrent)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +  
  labs(
    title = "Proportion of Concurrent HIV/AIDS Diagnoses by Neighborhood",
    x = "Neighborhood",
    y = "Proportion of Concurrent Diagnoses"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 8),
    plot.title = element_text(hjust = 0.5, size = 14)
  )
```

---------------------------------

### Pattern Discussion

The neighborhoods are ranked by their proportions, with Greenwich Village-Downtown having the lowest proportion and Sheepshead Bay the highest. The visualization highlights geographic variability in concurrent diagnoses, suggesting that certain neighborhoods may face distinct challenges in early detection or healthcare access.