---
title: "Exploratory Data Analysis"
author: "Tong Su, Shike Zhang"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library(tidyverse)
library(plotly)
library(dplyr)
library(ggplot2)
```

```{r, include=FALSE}
knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

---------------------------------

# Data Import and Cleaning

---------------------------------

```{r}
# Import the HIV and UHF datasets
HIV_df =
  read.csv("./data/HIV_AIDS_Diagnoses_by_Neighborhood__Sex__and_Race_Ethnicity_20241118.csv")|>
  janitor::clean_names()

UHF_df = 
  read.csv("./data/UHF_Neighborhoods.csv")|>
   janitor::clean_names()

# Clean HIV dataset
HIV_cleaned_df = HIV_df |>
  # Select rows where year 2016~2021
  filter(year >= 2016) |>
  # Convert Borough, Neighborhood (U.H.F), SEX, RACE/ETHNICITY to factors
  mutate(
    borough = as.factor(borough),
    neighborhood_u_h_f = as.factor(neighborhood_u_h_f),
    sex = as.factor(sex),
    race_ethnicity = as.factor(race_ethnicity)
  ) %>%
  # Convert necessary columns to numeric, handling non-numeric values
  mutate(
    total_number_of_hiv_diagnoses = as.numeric(total_number_of_hiv_diagnoses),
    hiv_diagnoses_per_100_000_population = as.numeric(hiv_diagnoses_per_100_000_population),
    total_number_of_concurrent_hiv_aids_diagnoses = as.numeric(total_number_of_concurrent_hiv_aids_diagnoses),
    proportion_of_concurrent_hiv_aids_diagnoses_among_all_hiv_diagnoses = as.numeric(proportion_of_concurrent_hiv_aids_diagnoses_among_all_hiv_diagnoses),
    total_number_of_aids_diagnoses = as.numeric(total_number_of_aids_diagnoses),
    aids_diagnoses_per_100_000_population = as.numeric(aids_diagnoses_per_100_000_population)
  )

# Merge datasets
HIV_UHF_df = HIV_cleaned_df |>
  inner_join(UHF_df, by = c("borough", "neighborhood_u_h_f" = "uhf_neighborhood"))
```

```{r, include=FALSE}
# Export the merged data
write.csv(HIV_UHF_df, "Merged_HIV_UHF_Data.csv", row.names = FALSE)
```

---------------------------------

## Data Dictionary

---------------------------------

* YEAR: Year of diagnosis.

* Borough: Borough information (some missing values).

* Neighborhood (U.H.F): Specific neighborhood identifiers.

* SEX: Sex of individuals.

* RACE/ETHNICITY: Racial/ethnic group classifications.

* TOTAL NUMBER OF HIV DIAGNOSES: Counts of HIV diagnoses.

* HIV DIAGNOSES PER 100,000 POPULATION: Rate of diagnoses normalized by population.

* TOTAL NUMBER OF CONCURRENT HIV/AIDS DIAGNOSES: Concurrent diagnoses of HIV and AIDS.

* PROPORTION OF CONCURRENT HIV/AIDS DIAGNOSES: Percentage of concurrent diagnoses.

* TOTAL NUMBER OF AIDS DIAGNOSES: Counts of AIDS diagnoses.

* AIDS DIAGNOSES PER 100,000 POPULATION: Rate of AIDS diagnoses normalized by population.

---------------------------------

## Data Summary

---------------------------------

```{r, echo=FALSE}
skimr::skim(HIV_df)
```

---------------------------------

# EDA

---------------------------------

## Temporal trends 

---------------------------------

### Temporal trends of HIV diagnoses
```{r, message=FALSE}
# Group by Borough and Year and calculate average HIV diagnosis rates
avg_hiv = HIV_UHF_df |>
  group_by(borough, year) |>
  summarise(Avg_HIV_Rate = mean(hiv_diagnoses_per_100_000_population, na.rm = TRUE))

avg_hiv
```

```{r}
# Plot HIV rates by Borough over time
ggplot(avg_hiv, aes(x = year, y = Avg_HIV_Rate, color = borough)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1.5) +
  labs(
    title = "HIV Diagnosis Rates by Borough Over Time",
    x = "Year",
    y = "Average HIV Diagnosis Rate (Per 100,000 Population)",
    color = "Borough"
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 10),
    legend.position = "bottom"
  )
```

* The Bronx consistently has the highest HIV diagnosis rates, with a decline between 2016 and 2018 followed by a rebound that peaked in 2021, likely driven by socio-economic challenges and healthcare access issues. In contrast, Staten Island has the lowest rates, with a sharp decline from 2016 to 2017 before stabilizing, suggesting effective interventions or a smaller population base. While most boroughs show declining or stabilizing trends, the Bronx's increasing rates underscore significant disparities and the urgent need for targeted, borough-specific strategies.

---------------------------------

### Temporal trends of AID diagnoses
```{r}
# Group by Borough and Year and calculate average AIDS diagnosis rates
avg_aids = HIV_UHF_df |>
  group_by(borough, year) |>
  summarise(Avg_AIDS_Rate = mean(aids_diagnoses_per_100_000_population, na.rm = TRUE))

avg_aids
```


```{r}
# Temporal trends of AIDS diagnoses
ggplot(avg_aids, aes(x = year, y = Avg_AIDS_Rate, color = borough, group = borough)) +
  geom_line(size = 1) +
  geom_point() +
  labs(
    title = "Temporal Trends: AIDS Diagnosis Rates Across Boroughs",
    x = "Year",
    y = "Average AIDS Diagnosis Rate (Per 100,000 Population)",
    color = "Borough"
  ) +
  theme_minimal()


```

* The Bronx experienced a dramatic spike in HIV diagnoses in 2021, reversing a declining trend from 2016 to 2019. This sharp increase suggests systemic issues such as delayed diagnoses, treatment interruptions, or lack of timely access to antiretroviral therapy (ART). In contrast, Staten Island's HIV diagnosis rates remain consistently low, similar to Queens, with only a slight increase from 2020 to 2021, warranting further investigation despite remaining well below rates in the Bronx and Manhattan.

---------------------------------

## Demographic Disparities

---------------------------------

### HIV Diagnoses by Sex and Race/Ethnicity
```{r}
# Group by Sex and Ethnicity with average rate of HIV
avghiv_sex_ethinicity = HIV_UHF_df |>
  group_by(borough, year,sex,race_ethnicity) |>
  summarise(Avg_HIV_Rate = mean(hiv_diagnoses_per_100_000_population, na.rm = TRUE))

avghiv_sex_ethinicity
```

```{r}
# Demographic disparities in HIV diagnoses
ggplot(avghiv_sex_ethinicity, aes(x = sex, y = Avg_HIV_Rate, fill = race_ethnicity)) +
  geom_boxplot() +
  labs(
    title = "Demographic Disparities: HIV Diagnosis Rates by Sex and Race/Ethnicity",
    x = "Sex",
    y = "Average HIV Diagnosis Rate (Per 100,000 Population)",
    fill = "Race/Ethnicity"
  ) +
  theme_minimal()

```

* Black populations have the highest HIV diagnosis rates, followed by Latino/Hispanic populations, while Asian/Pacific Islander and White groups consistently show the lowest rates, highlighting stark racial disparities. Males across all racial/ethnic groups have higher diagnosis rates than females, particularly Black and Latino/Hispanic males, who show significant outliers, indicating areas requiring focused interventions. Although females have lower overall rates, Black females face notable disparities compared to other groups. These patterns underscore systemic inequities in access to prevention, testing, and treatment, with Black and Latino/Hispanic populations requiring prioritized healthcare services and interventions.

---------------------------------

## Concurrent Diagnoses

---------------------------------

### Proportion of concurrent diagnoses by neighborhood

```{r}
# Group by proportion of concurrent diagnoses by neighborhood
proprotion_diagnoses_neighborhood = HIV_UHF_df |>
  group_by(borough, year,neighborhood_u_h_f) |>
  summarise(Avg_proportion_Rate = mean(proportion_of_concurrent_hiv_aids_diagnoses_among_all_hiv_diagnoses, na.rm = TRUE)
            )

proprotion_diagnoses_neighborhood
```

```{r}
ggplot(proprotion_diagnoses_neighborhood, aes(x = neighborhood_u_h_f, y = Avg_proportion_Rate,fill=borough)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Proportion of Concurrent Diagnoses by Neighborhood",
    x = "Neighborhood",
    y = "Average Proportion of Concurrent Diagnoses (%)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size=6))
```

* Within the same borough, concurrent HIV/AIDS diagnosis rates vary widely across neighborhoods, with certain areas in Brooklyn (e.g., Bedford-Stuyvesant) and Manhattan (e.g., Greenwich Village, Chelsea) showing exceptionally high rates, often exceeding 150%, while Staten Island and Queens maintain lower percentages. These outliers in Brooklyn and Manhattan suggest structural challenges related to healthcare access, awareness, or socio-economic factors. Community-level disparities emphasize the role of localized healthcare infrastructure and socio-economic conditions. The high rate of concurrent diagnoses reflects late HIV detection, often when patients have progressed to AIDS, highlighting an urgent need for earlier testing and intervention efforts.

---------------------------------

### Proportion of Concurrent Diagnoses by  Sex

```{r}
# Group by proportion of concurrent diagnoses by borough and sex with average rate of HIV
proprotion_diagnoses_sex = HIV_UHF_df |>
  group_by(borough, year,sex) |>
  summarise(Avg_proportion_Rate = mean(proportion_of_concurrent_hiv_aids_diagnoses_among_all_hiv_diagnoses, na.rm = TRUE)
            )

proprotion_diagnoses_sex
```

```{r}
# Concurrent diagnoses by borough and sex
ggplot(proprotion_diagnoses_sex, aes(x = borough, y = as.numeric(Avg_proportion_Rate), fill = sex)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Proportion of Concurrent Diagnoses by Borough and Sex",
    x = "Borough",
    y = "AverageProportion of Concurrent Diagnoses (%)",
    fill = "Sex"
  ) +
  theme_minimal()

```

* The proportion of concurrent diagnoses varies widely across boroughs, with Staten Island having the highest overall rate, particularly among females, which stands out as a significant outlier warranting further investigation. Queens also shows a relatively high proportion among females, while other boroughs display a more balanced distribution. The Bronx, despite its high HIV prevalence, demonstrates relatively balanced proportions, suggesting effective early detection and treatment efforts. Most boroughs exhibit higher rates among males, aligning with data on delayed diagnoses among men, especially men who have sex with men (MSM).

---------------------------------

### Proportion of Concurrent Diagnoses by Race/Ethnicity

```{r}
# Group by proportion of concurrent diagnoses by borough and ethnicity with average rate of HIV
proprotion_diagnoses_race = HIV_UHF_df |>
  group_by(borough, year,race_ethnicity) |>
  summarise(Avg_proportion_Rate = mean(proportion_of_concurrent_hiv_aids_diagnoses_among_all_hiv_diagnoses, na.rm = TRUE)
            )

proprotion_diagnoses_race
```


```{r}
# Concurrent diagnoses by race/ethnicity and borough
ggplot(proprotion_diagnoses_race, aes(x = race_ethnicity, y = as.numeric(Avg_proportion_Rate), fill = borough)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Proportion of Concurrent Diagnoses by Race/Ethnicity and Borough",
    x = "Race/Ethnicity",
    y = "Average Proportion of Concurrent Diagnoses (%)",
    fill = "Borough"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

* Asian/Pacific Islanders in the Bronx have the highest rates of concurrent diagnoses, surpassing all other racial/ethnic groups across boroughs. Staten Island also exhibits very high rates among Whites and the Other/Unknown category, with similar trends in Queens. Black and Latino/Hispanic populations show consistent proportions across boroughs but remain highest in the Bronx. Brooklyn and Queens have notably lower percentages of Asian/Pacific Islanders compared to the Bronx. High rates in specific groups in the Bronx and Staten Island likely reflect delayed diagnoses or systemic barriers such as limited access to care, stigma, or socioeconomic challenges. Overrepresentation of Other/Unknown racial groups suggests potential data limitations or gaps.


